### 2014-04-19 text/Chapter03.tex
Semantics of structural liveness (Chapter 3)

\paragraph{Structural typing rules.}

The rules are split into two groups. We first discuss the standard rules.

is good

@@

\begin{figure}[t]
{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{const}
  {c:\tau \in \Delta}
  {\coctx{()}{\alift{}} \vdash c : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x : \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\vspace{0.5em}

{\small b.) Structural rules for context manipulation}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{0}} \vdash e : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
\end{equation*}

@@

>>>>

@@

\begin{figure}[t]
{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{const}
  {c:\tau \in \Delta}
  {\coctx{()}{\alift{}} \vdash c : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\vspace{0.5em}

{\small b.) Structural rules for context manipulation}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\ident{D}} }} \vdash e : \sigma} 
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
  {\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
\quad
\begin{array}{l}
\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}
\begin{equation*}
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}\sqcap\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
~
\begin{array}{l}
\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}

@@

\paragraph{Structural typing rules.}
The structural typing rules are shown in Figure~\ref{fig:applications-struct-live} (b). They mirror
the rules know from sub-structural type systems (Section~\ref{sec:path-logic}). Weakening (\emph{weak})
extends the context with a single unused variable $x$ and adds the $\ident{D}$ annotation to the
vector of coeffects.

The variable is always added to the end as in the (\emph{abs}) rule. However, the exchange rule 
(\emph{exch}) lets us arbitrarily reorder variables. It flips the variables $x$ and $x'$ and their
corresponding coeffect annotations in the vector. This is done by requiring that the lengths of the
remaining, unchanged, parts of the vectors match.

Finally, contraction (\emph{contr}) makes it possible to use a single variable multiple times.
Given a judgement that contains variables $y$ and $z$, we can derive a judgement for an expression
where both $z$ and $y$ are replaced by a single variable $x$. Their annotations $\cclrd{s}, \cclrd{t}$
are combined into $\cclrd{s} \sqcap \cclrd{t}$, which means that $x$ is live if either $z$ or $y$
were live in the original expression.

\paragraph{Example.} To demonstrate how the system works, we consider the expression
$(\lambda x \rightarrow v)~y$. This is similar to an example where flat liveness mistakenly marks
the entire context as live. Despite the fact that the variable $y$ is accessed (syntactically), it
is not live -- because the function that takes it as an argument always returns $v$.

The typing derivation for the body uses (\emph{var}) and (\emph{abs}). However, we also need (\emph{weak})
to add the unused variable $x$ to the context:
%
\begin{equation*}
\tyrule{weak}
{ \tyruler{var}
    {}
    { \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash v : \tau } }
{ \tyruler{abs}
    { \coctx{v\!:\!\tau, x\!:\!\tau}{\alift{ \cclrd{\ident{L}},\cclrd{\ident{D}} }} \vdash v : \tau }
    { \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash (\lambda x \rightarrow v) : \tau \xrightarrow{\cclrd{\ident{D}}} \tau }}
\end{equation*}
%
The interesting part is the use of the (\emph{app}) rule in the next step. Although the variable $y$ is live in the expression $y$,
it is marked as dead in the overall expression, because the function is annotated with $\ident{\cclrd{D}}$:
%
\begin{equation*}
\hspace{-2em}
\tyrule{app}
  {
    \begin{array}{l}
    \vspace{-1.2em}
    \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash (\lambda x \rightarrow v) : \tau \xrightarrow{\cclrd{\ident{D}}} \tau 
    \end{array} &
    \tyruler{var}{}{\coctx{y\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash y : \tau}
  }
  { 
  \inference
  	{ \coctx{v\!:\!\tau, y\!:\!\tau}{\alift{ \cclrd{\ident{L}} } \cons\, ({ \cclrd{\ident{D}} \sqcup \alift{\cclrd{\ident{L}}} })} 
  	     \vdash (\lambda x \rightarrow v)~y : \tau }
  	{ \coctx{v\!:\!\tau, y\!:\!\tau}{\alift{ \cclrd{\ident{L}}, \cclrd{\ident{D}} }} \vdash (\lambda x \rightarrow v)~y : \tau }  	
  }
\end{equation*}
%
The application is written in two steps -- the first one directly applies the (\emph{app}) rule
and the second one simplifies the coeffect annotation. The key part is the use of the scalar-vector 
operator $\cclrd{\ident{D}} \sqcup \alift{\cclrd{\ident{L}}}$. Using the definition of the scalar-vector
extension, this equals $\alift{\cclrd{\ident{D}} \sqcup \cclrd{\ident{L}}}$ which is
$\alift{\cclrd{\ident{D}}}$.

\begin{figure}[t]
{\small a.) Semantics of ordinary expressions}
\begin{equation*}
\tag{\emph{var}}
\sem{\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} = \lambda(x) \rightarrow x
\end{equation*}
\vspace{-1.5em}
\begin{equation*}
\tag{\emph{const}}
\sem{\coctx{()}{\alift{}} \vdash c : \tau} = \lambda () \rightarrow \delta(c)
\end{equation*}
\vspace{-1.5em}
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{abs}}
\begin{array}{l}
\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda y . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} = \lambda \textbf{v} \rightarrow\\[-0.25em]
\qquad\lambda y \rightarrow \sem{\coctx{\Gamma, y \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}~(\textbf{v}, y)
\end{array}
\end{equation*}
\vspace{-1.0em}
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{app-1}}
\begin{array}{l}
\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{L}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
\end{array}
\end{equation*}
\vspace{-1.0em}
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{app-2}}
\begin{array}{l}
\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{D}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 }~\mathbf{v_1}~
  \kvd{in}~g~()
\end{array}
\end{equation*}
\vspace{0.5em}

{\small b.) Semantics of structural context manipulation}
\begin{equation*}
\tag{\emph{weak}}
\sem{\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\ident{D}} }} \vdash e : \sigma} = 
  \lambda(\textbf{v}, ()) \rightarrow
    \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}~\textbf{v}
\end{equation*}
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{exch}}
\begin{array}{l}
\sem{\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} = 
  \lambda(\mathbf{v_1}, y, x, \mathbf{v_2}) \rightarrow\\[-0.25em]
\qquad
  \sem{\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~
  (\mathbf{v_1}, x, y, \mathbf{v_2})
\end{array}
\end{equation*}
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{contr-1}}
\begin{array}{l}
\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, (), \mathbf{v_2}) \rightarrow \\[-0.25em]
\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} ~(\mathbf{v_1}, (), (), \mathbf{v_2})
\end{array}
\end{equation*}
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{contr-2}}
\begin{array}{l}
\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, x, \mathbf{v_2}) \rightarrow \\[-0.25em]
\quad\begin{cases}
~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, x, \mathbf{v_2})\\[-0.25em]
~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, (), x, \mathbf{v_2})\\[-0.25em]
~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, (), \mathbf{v_2})
\end{cases}
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Semantics of structural liveness}
\label{fig:applications-struct-livesem}
\end{figure}
\vspace{-0.5em}

\paragraph{Semantics.}
When defining the semantics of flat liveness calculus, we used an indexed form of the option type
$1 + \tau$ (which is $1$ for dead contexts and $\tau$ for live contexts). In the semantics of 
expressions, the type wrapped the entire context, \ie~$1+(\tau_1 \times \ldots \times \tau_n)$.
In the structural version, the semantics wraps individual elements of the free-variable context
pair: $(1+\tau_1) \times \ldots \times (1+ \tau_n)$. For each variable, the type is indexed by 
the corresponding annotation. More formally:
%
\begin{equation*}
\begin{array}{r}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
  ~:~ (\tau'_1 \times \ldots \times \tau'_n) \rightarrow \tau\\[0.5em]
\textnormal{where}~\tau'_i = \begin{cases}
\tau_i & (\cclrd{r_i} = \cclrd{\ident{L}})\\
1      & (\cclrd{r_i} = \cclrd{\ident{D}})
\end{cases}
\end{array}
\end{equation*}
%
Note that the product of the free variables is not an ordinary tuple of our language, but a special
construction. This follows from the asymmetry of $\lambda$-calculus, as discussed in 
Section~\ref{sec:applications-strucutre-vec}. Functions take just a single input and so they are
interpreted in the same way as in flat calculus:
%
\begin{equation*}
\sem{\tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2 \qquad\qquad
\sem{\tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
\end{equation*}
%
The rules that define the semantics are shown in Figure~\ref{fig:applications-struct-livesem}.
To make the definition simpler, we are somewhat vague when working with products. We write 
variables of product type such as $\mathbf{v}$ in bold-face and individual values like $x$ in 
normal face. We freely re-associate products and so $(\mathbf{v}, x)$ should not be seen as a
nested product, but simply a product with a number of variables represented as another product
$\mathbf{v}$ with one more variable $x$ at the end. We shall be more precise in Chapter~\ref{ch:structural}.

In (\emph{var}), the context contains just a single variable and so we do not even need to apply
projection; (\emph{cosnt}) receives no variables and uses global constant lookup function $\delta$.
In (\emph{abs}), we obtain two parts of the context and combine them into $(\mathbf{v}, x)$. This
works the same way regardless of whether the variables are live or dead. For simplicity, we omit
sub-coeffecting, which just turns some of the available values $v_i$ to unit values $()$.

In application, we again need to ``implement'' dead code elimination. When the input parameter
of the function $g$ is live (\emph{app-1}), we first evaluate $e_2$ and then pass the result to
$g$. When the parameter is dead (\emph{app-2}), we do not need to evaluate $e_2$ and so all values
in $\mathbf{v_2}$ can be dead, \ie~$()$.

In the structural rules, (\emph{weak}) receives context containing a dead variable as the last one.
It drops the $()$ value and evaluates the expression in a context $\mathbf{v}$. Exchange (\emph{exch})
simply swaps two variables. In contraction, we either duplicate a dead value (\emph{contr-1}),
or a live value (\emph{contr-2}). In the latter, one of the duplicates may be dead and 
so we need to consider three separate cases.

\paragraph{Summary.}
The structural coeffect system for tracking liveness analysis

semantics is ``the other way round''
